### 2013-07-29 doc/asplos14/communication.tex

Each processing core in a given system node is responsible for
aggregating and sending the resulting messages from all processors on
that node to a set of destination nodes.  Each core periodically
executes a task responsible for sending messages.  This task examines
the private (to each core) message lists for each destination node it
is responsible for managing and, if the list is long enough or a
message has waited past a time-out period, all messages to a given
destination system node from that source system node are sent.
Aggregating and sending a message involves manipulating a set of
shared data-structures (the message lists). This is done using CAS
(compare-and-swap) operations to avoid high synchronization costs.
Note that we use a per-processor array of message lists
that is only periodically modified across processor cores because we
empirically determined this approach was faster (sometimes
significantly) than a global per-system node array of message lists.

Each node has a region of memory with send buffers where the final aggregated
messages are built. These buffers are visible to the network card, and
messages can be sent with user-mode operations only. When the worker
responsible for outbound messages to a given system node has received a
sufficient number of message send requests or a timeout is reached, the
linked list of messages is walked and messages are copied to a send buffer.
This process requires careful prefetching because most of the outbound
messages are \emph{not\/} in the processor cache at this time (recall that a
core can be aggregating messages originating from other cores in the same
node). Once the send buffer has been formed, it is handed off to GASNet for
transfer to the remote system node. RDMA is used if the underlying network
supports it.

@@

Once the remote system node has received the message buffer, a
management task is spawned to manage the unpacking process.  The
management task spawns a task on each processor at the receiving
system to simultaneously unpack messages destined for that processor.
Upon completion, these unpacking tasks synchronize with the management
task.  Once all cores have processed the message buffer, the management
task sends a reply to the sending system node indicating the
successful delivery of the messages.

>>>>

Each processing core in a given system node is responsible for
aggregating and sending the resulting messages from all cores on
that node to a set of destination nodes.  Each core periodically
executes a task responsible for sending messages.  This task examines
the private (to each core) message lists for each destination node it
is responsible for managing and, if the list is long enough or a
message has waited past a time-out period, all messages to a given
destination system node from that source system node are sent.
Aggregating and sending a message involves manipulating a set of
shared data-structures (the message lists). This is done using CAS
(compare-and-swap) operations to avoid high synchronization costs.
Note that we use a per-core array of message lists
that is only periodically modified across processor cores after
experimentally determining that this approach was faster (sometimes
significantly) than a global per-system node array of message lists.

Each node has a region of memory with send buffers where the final aggregated
messages are built. These buffers are visible to the network card, and
messages are sent with user-mode operations only. When the worker
responsible for outbound messages to a given system node has received a
sufficient number of message send requests or a timeout is reached, the
linked list of messages is walked and messages are copied to a send buffer.
This process requires careful prefetching because most of the outbound
messages are \emph{not\/} in the processor cache at this time (recall that a
core can be aggregating messages originating from other cores in the same
node). Once the send buffer has been formed, it is handed off to GASNet for
transfer to the remote system node. RDMA is used if the underlying network
supports it.

@@

Once the remote system node has received the message buffer, a
management task is spawned to manage the unpacking process.  The
management task spawns a task on each core at the receiving
system to simultaneously unpack messages destined for that core.
Upon completion, these unpacking tasks synchronize with the management
task.  Once all cores have processed the message buffer, the management
task sends a reply to the sending system node indicating the
successful delivery of the messages.

### 2013-07-29 doc/asplos14/intro.tex
abstract intro numbers

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low temporal and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on effective data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

@@

We believe we can support irregular applications with good performance and cost effectively with commodity hardware for two main reasons. First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth. Second, commodity networks offer high bandwidth as long as messages are large enough. We build on these two observations and develop \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications. \Grappa exploits fast processors and the memory hierarchy to provide a lightweight user-level tasking layer that supports a context switch in as little as 38ns and can sustain a large number of active workers. It bridges the commodity network bandwidth gap with a communication layer that combines short messages originating from many concurrent workers into larger packets.

@@

Our evaluation of \Grappa shows that its core components provide scalable performance for irregular applications. It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.  The communication layer helps \Grappa achieve over 1 billion random updates per second across 64 system nodes. \Grappa provides performance approaching and in some cases exceeding that of the custom-hardware XMT, as well as MPI and UPC implementations of the same benchmarks. When we compare \Grappa to the XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark. When compared with MPI and UPC, \Grappa is between 11X faster and 5.4X slower depending on the benchmark. We find that the benchmarks that are faster contain specialized implementations of features \Grappa provides generally.

>>>>

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low temporal and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

@@

We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons. First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth. Second, commodity networks offer high bandwidth as long as messages are large enough. We build on these two observations and develop \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications. \Grappa exploits fast processors and the memory hierarchy to provide a lightweight user-level tasking layer that supports a context switch in as little as 38ns and can sustain a large number of active workers. It bridges the commodity network bandwidth gap with a communication layer that combines short messages originating from many concurrent workers into larger packets.

@@

Our evaluation of \Grappa shows that its core components provide scalable performance for irregular applications. It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.  The communication layer helps \Grappa achieve over 1 billion random updates per second across 64 system nodes. \Grappa provides performance approaching and in some cases exceeding that of the custom-hardware XMT, as well as MPI and UPC implementations of the same benchmarks. When we compare \Grappa to the XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark. When compared with MPI, \Grappa is 9X faster on a random access microbenchmark and 3.5X to 5.4X slower on applications. We find that the benchmarks that are faster contain specialized implementations of features \Grappa provides generally.

### 2013-07-29 doc/asplos14/tasks.tex

\Grappa provides three methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second,
the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
small non-blocking piece of code on a particular core in the system without waiting for
 a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} routine for this purpose.

\begin{figure}[htbp]
  \begin{center}
	\begin{description}
    \item[\texttt{spawn( Functor f )}] \hfill \\
      Creates a new stealable task
    \item[\texttt{parallel\_for( start, end, Functor iteration )}] \hfill \\
      Executes iterations of a loop as stealable tasks that
      take the iteration index as an argument
    \item[\texttt{call\_on( core, Functor f )}] \hfill \\
      Runs a limited function on a specific core without
      consuming a stack or context switching
	\end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
      } % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

>>>>

\Grappa provides three methods for expressing parallelism. First, a single task can be created to execute in parallel with the current task by calling \texttt{spawn} with a functor. This adds it to the queue of tasks which will be executed the next time a worker is available.
Second,
the programmer can invoke a parallel for loop with \texttt{parallel\_for}, provided that the trip count is
known at loop entry. The programmer specifies a functor which takes the loop index as a parameter, and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold.
Third, parallelism can be expressed via asynchronous delegate operations, which are explained next, in Section~\ref{sec:memory}.
% Third, a programmer may want to execute an active message~\cite{vonEicken92}; that is, to run a
% small non-blocking piece of code on a particular core in the system without waiting for
 % a free worker to become available. For example, custom synchronization primitives are implemented as an active message executed on the core where the data
% lives. \Grappa provides the \texttt{call\_on} routine for this purpose.

% \begin{figure}[htbp]
%   \begin{center}
%   \begin{description}
%     \item[\texttt{spawn( Functor f )}] \hfill \\
%       Creates a new stealable task
%     \item[\texttt{parallel\_for( start, end, Functor iteration )}] \hfill \\
%       Executes iterations of a loop as stealable tasks that
%       take the iteration index as an argument
%     \item[\texttt{call\_on( core, Functor f )}] \hfill \\
%       Runs a limited function on a specific core without
%       consuming a stack or context switching
%   \end{description}
%     \begin{minipage}{0.95\columnwidth}
%       \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
%       } % \vspace{-4ex}}
%     \end{minipage}
%     %\vspace{-3ex}
%   \end{center}
% \end{figure}

### 2013-07-29 doc/sosp13/memory.tex

Delegate operations can be either {\em synchronous} or {\em asynchronous}.
With synchronous operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in \Grappa
is implemented using synchronous delegate operations. Often remote data
accesses do not need to be performed synchronously. In order to avoid
unnecessary waiting, we support asynchronous delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by asynchronous delegate
invocations. Delegate write operations may also be performed asynchronously,
and since they do not return data, they need a mechanism to detect completion.
\Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
outstanding asynchronous operations of the local core. The code and can probe
the state of this object to determine when all outstanding requests are
complete. This is necessary to, say, make sure all outstanding operations are
complete before executing a synchronization operation.

>>>>

Delegate operations can be either {\em synchronous} or {\em asynchronous}.
With synchronous operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in \Grappa
is implemented using synchronous delegate operations. On the other hand, remote data
accesses often do not need to be performed synchronously. In order to avoid
unnecessary waiting, we support asynchronous delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by asynchronous delegate
invocations. Delegate write operations may also be performed asynchronously,
and since they do not return data, a mechanism to detect completion is needed.
\Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
outstanding asynchronous operations of the local core. Threads that depend on completion may block on this object
the state of this object, and they will be woken when all outstanding requests are
complete.
